An Object-Logic Explanation for the Inconsistency in G\"odel's Ontological Theory

### 2016-07-09 Papers/2016/KI/main.tex
Improve introduction

\title{The Inconsistency in G\"odel's Ontological Argument: A Success
  Story for Artificial Intelligence in Metaphysics (Extended Abstract)}
\author{Christoph Benzm\"uller\inst{1}\thanks{This work has been supported by
    the German Research Foundation DFG under grants BE2501/9-1,2} \and Bruno Woltzenlogel Paleo\inst{2} 
}

@@

\begin{abstract}
  We discusses the discovery of the inconsistency in G\"odel's
  ontological argument as a success story for artificial
  intelligence. Despite the popularity of G\"odel's argument this
  inconsistency remained unnoticed until 2013, when it was detected
  automatically by the higher-order theorem prover
  \textsc{Leo-II}.  % Understanding and verifying the
  % refutation generated by the prover turned out to be a time-consuming
  % task. Its completion, as reported here, required the
  % reconstruction of the refutation in the Isabelle proof assistant.% , and it
  % also led to a novel and more efficient way of automating
  % higher-order modal logic \textbf{S5} with a universal accessibility
  % relation. Furthermore, the development of an improved syntactical
  % hiding for the utilized logic embedding technique allows the refutation to be
  % presented in a human-friendly way, suitable for non-experts in the
  % technicalities of higher-order theorem proving. This brings us a
  % step closer to wider adoption of logic-based artificial intelligence
  % tools by philosophers.

  (This text is a significantly shortened version of our contribution to IJCAI 2016.)
\end{abstract}

@@

 % In philosophy lectures at universities the
% argument is regularly presented as a masterpiece argument in
% metaphysics. Since 2013, when we~\cite{J30,C40} first
% reported our successful initial computer-assisted
% analysis of G\"odel's proof and Scott's variant,
% our work has received a media repercussion on a global scale\footnote{A
%   collection of news articles is available at {\scriptsize
%     \url{https://github.com/FormalTheology/GoedelGod/blob/master/Press/LinksToNews.md}}},
% and numerous bloggers commented on the proof
% \cite{fuhrmann15:_blogg_goedel}.

However, G\"odel's original version of the axioms used in his
manuscript \cite{GoedelNotes} are inconsistent, and this fact has
remained unnoticed to philosophers until 2013, when it was detected
automatically by the higher-order theorem prover \textsc{Leo-II}
\cite{C40}.  We here report on the extraction, reconstruction and
verification of an informal, human intuitive argument for this
inconsistency in the Isabelle/HOL proof assistant.  The respective
presentation
of this intuitive argument further extends and improves our contribution 
to IJCAI 2016 \cite{C55}.

% The verified refutation (discussed
% in \S\ref{sec:inconsistency}) displays a surprisingly accessible
% explanation of the inconsistency, which is philosophically profound
% and never presented in the literature. 
% The detection of this
% inconsistency in combination with the work reported here thus
% demonstrates that higher-order automated theorem provers are capable
% of assisting in the discovery and elucidation of \emph{new} and
% philosophically relevant knowledge.

% On the technical side, the quest for constructing a compelling
% refutation, capable of convincing also human non-experts, led to an
% improvement of the syntax of the embedding of modal logics in Isabelle/HOL
% (as discussed in \S\ref{sec:improvedsyntax}). With the new syntax, a
% (nearly) perfect match between the original pen and paper
% presentations and our encoding in Isabelle/HOL is feasible. A more
% user-friendly syntax, as reported here, is clearly an important
% prerequisite for promoting the theorem proving technology employed
% here to a wider community of philosophers, who are not necessarily
% experts in automated reasoning or HOL.

% Another novel contribution reported here (in
% \S\ref{sec:improvedembedding}) is the implementation of an alternative
% embedding for the 
% % slightly stronger 
% more effective modal logic \SFiveU, which is
% based on a universal accessibility relation. Our experiments have
% shown that the new embedding is more efficient, as the following two
% previously open problems can now be solved:
% \begin{itemize}
% \item Automatically proving the final theorem T3 (\textit{Necessarily, there
%   exists God}), directly from Scott's \cite{ScottNotes} (consistent) axioms
%   alone, without relying on the argument's intermediate argumentation
%   steps (i.e., lemmata).
% \item Automatically verifying, in Isabelle/HOL, the proof of the modal
%   collapse \cite{Sobel}, which is one of the most strongly criticized
%   logical consequences of the argument's axioms.
% \end{itemize}

In our experiments we have employed theorem provers for classical
higher-order logic (HOL; cf.~\cite{andrewsSEP} and the references
therein), even though G\"odel's proof is actually formulated in
higher-order \emph{modal} logic (HOML; cf.~\cite{homl} and the
references therein). To bridge between the two logics we adopt the
logic embedding approach \cite{J23,C40}.

%\paragraph{Related Work.}
Applications of (first-order) theorem proving technology in
metaphysics were first reported by Fitelson, Oppenheimer and
Zalta~\cite{FitelsonZalta,oppenheimer11}. Zalta also coined the term
\textit{Computational Metaphysics}. Later on, Rushby~\cite{rushby13} used the proof
assistant \textsc{PVS}. Common to both works is a
significant amount of proof-hand-coding work as well as their focus on
a non-modal formalization of St. Anselm's simpler
and older ontological argument. % In contrast, the greater complexity of
% G\"odel's argument requires the formalization and automation of
% variants of \emph{higher-order} and \emph{modal} logics.

\begin{figure}\centering 
  \includegraphics[width=.495\textwidth]{./Scott_S5U.png}
  \includegraphics[width=.495\textwidth]{./InconsistencyIsabelleK.png}
  % \caption{Full Automation of T3 in \SFiveU; Consistency of Scott's
  % Axioms;  Automatic Verification of Modal Collapse} 
\caption{Scott's consistent axioms (left) and proof of the
  inconsistency of (a subset of) G\"odel's original  axioms (right)}
  % Axioms;  Automatic Verification of Modal Collapse} 
\label{Scott_Goedel}
\end{figure} 
\paragraph{A Brief History.}\label{sec:history}
% St. Anselm's ontological argument \cite{Proslogion} can be regarded as
% the ancestor of modern ontological arguments such as G\"odel's. In the
% millenium between Anselm and G\"odel, many philosophers modified and
% arguably improved Anselm's argument. Of particular importance to
% G\"odel was the work of Leibniz~\cite{Adams}.  
% Although G\"odel's notion of positive property is not exactly the same
% as Leibniz's notion of perfection, 
G\"odel's manuscript can be considered a translation of Leibniz's
ideas on the argument into modern modal logic. G\"odel
discussed his manuscript with Scott, who shared a slightly different
version with a larger public. Scott's version of the axioms and
definitions, formalized in Isabelle, is shown in
Fig.~\ref{Scott_Goedel}. 
The main difference to G\"odel's version is an
extra conjunct in the definition of \emph{essence} (\emph{ess}); cf. Fig.~\ref{Scott_Goedel}. For Scott,
an essential property of an individual must be possessed by
him/her. For G\"odel, this is not required.

% \begin{wrapfigure}[20]{r}{0.55\textwidth}
%     \centering \vskip-1em
%     \centerline{\includegraphics[width=0.52\textwidth]{./InconsistencyIsabelleK.png}}
%  \caption{Inconsistency of G\"odel's Axioms in HOML \K verified in Isabelle/HOL} \label{InconsistencyIsabelleK}
% \end{wrapfigure}
G\"odel's omission has been
considered inessential and merely an oversight by many. 
% For instance,
% Hazen [\cite[p.365]{Hazen}] states that ``G\"odel left this
% clause out [\ldots] but this appears to have been an oversight''.
% -- it is
%included in related manuscripts''. 
For more than four decades, its serious consequences remained unnoticed, 
despite numerous analysis and criticisms of the
argument.%  Especially since the discovery by Sobel
% \cite{Sobel} that modal collapse (MC)\footnote{The modal collapse,
%   $\phi\rightarrow \Box \phi$, states that contingent truth implies
%   necessary truth; it can be interpreted as \textit{everything is
%     pre-determined} or even \textit{there is no free will}. } is
% entailed by G\"odel's (or also Scott's) axioms, several variants have
% been proposed
% \cite{Anderson,AndersonGettings,Hajek1,Hajek2,Hajek3,Bjordal}
% attempting to avoid the modal collapse. Many of these variants omit
% the crucial conjunct in the definition of essence as well.\footnote{As
%   these variants also change other axioms, on which the inconsistency
%   of G\"odel's axioms depends, it is not necessarily the case that
%   these variants are also inconsistent; they must be analyzed
%   separately.} Opponents of the argument
% (e.g. Oppy [\cite[p.226--227]{oppy96:_goedel_ontol_argum};
% \cite[p.364]{oppy00:_respon_gettin};
% \cite[p.1068]{oppy08:_higher_order_ontol_argum}]) have also proposed
% parodies and other criticisms, referring to variants where the
% conjunct is omitted.

>>>>

\title{An Object-Logic Explanation for the Inconsistency in G\"odel's Ontological Theory}
\author{Christoph Benzm\"uller\inst{1}\thanks{This work has been supported by
    the German Research Foundation DFG under grants BE2501/9-1,2} \and Bruno Woltzenlogel Paleo\inst{2} 
}

@@

\begin{abstract}
  This paper discusses the discovery of the 
  inconsistency in G\"odel's
  ontological argument as a success story for artificial
  intelligence. Despite the popularity of G\"odel's argument, this
  inconsistency remained unnoticed until 2013, 
  when it was detected automatically by the 
  higher-order theorem prover \textsc{Leo-II} \cite{ECAI2014}. Complementing the meta-logic explanation for the inconsistency available in our IJCAI 2016 paper \cite{C55}, we present here a new purely object-logic explanation that does not rely on semantic argumentation.

  % (This extended abstract is mostly a shortened version of our IJCAI 2016 paper \cite{C55}, but the explanation of the inconsistency is more detailed here.)
\end{abstract}

% ToDo: Use \textsc for all system names

@@

Scott's version of G\"odel's argument was automatically reconstructed by the higher-order theorem prover \textsc{Leo-II} \cite{C40} and its correctness was verified step-by-step in the Coq proof assistant \cite{CSR}. To bridge the gap between higher-order logics (HOL; cf.~\cite{andrewsSEP} and the references
therein), as used by these systems, and higher-order \emph{modal} logics (HOML; cf.~\cite{homl} and the
references therein), on which the ontological argument argument, the logic embedding approach \cite{J23,C40} was used.

However, G\"odel's original axioms, as used in his
manuscript \cite{GoedelNotes}, are inconsistent. This fact has
remained unnoticed to philosophers until 2013, 
when \textsc{Leo-II} found a surprising refutation of the axioms.

In \cite{C55}, we have extracted from Leo-II's machine-oriented refutation an informal and human-oriented intuitive explanation for the inconsistency, and we have reconstructed and
verified it in the Isabelle proof assistant. However, that explanation relied on reasoning at the meta-logic (HOL) level, which was only possible because of the embedding. Here we complement that work with a purely object-logic (HOML) explanation.

Applications of (first-order) theorem proving technology in
metaphysics were first reported by Fitelson, Oppenheimer and
Zalta~\cite{FitelsonZalta,oppenheimer11}. 
%Zalta also coined the term
%\textit{Computational Metaphysics}. 
Later on, Rushby~\cite{rushby13} used the \textsc{PVS} proof
assistant. Common to both works is a
significant amount of proof-hand-coding work as well as their focus on
a non-modal formalization of St. Anselm's simpler
and older ontological argument.

\begin{figure}\centering 
  \includegraphics[width=.495\textwidth]{./Scott_S5U.png}
  \includegraphics[width=.495\textwidth]{./InconsistencyIsabelleK.png}
  % \caption{Full Automation of T3 in \SFiveU; Consistency of Scott's
  % Axioms;  Automatic Verification of Modal Collapse} 
\caption{Scott's consistent axioms (left) and proof of the
  inconsistency of (a subset of) G\"odel's original  axioms (right)}
  % Axioms;  Automatic Verification of Modal Collapse} 
\label{Scott_Goedel}
\end{figure}

\paragraph{A Brief History.}\label{sec:history}
% St. Anselm's ontological argument \cite{Proslogion} can be regarded as
% the ancestor of modern ontological arguments such as G\"odel's. In the
% millenium between Anselm and G\"odel, many philosophers modified and
% arguably improved Anselm's argument. Of particular importance to
% G\"odel was the work of Leibniz~\cite{Adams}.  
% Although G\"odel's notion of positive property is not exactly the same
% as Leibniz's notion of perfection, 
G\"odel's manuscript can be considered a translation of Leibniz's
ideas on the argument into modern modal logic. G\"odel
discussed his manuscript with Scott, who shared a slightly different
version with a larger public. Scott's version of the axioms and
definitions, formalized in Isabelle, is shown in
Fig.~\ref{Scott_Goedel}. 
The main difference to G\"odel's version is an
extra conjunct in the definition of \emph{essence} (\emph{ess}); cf. Fig.~\ref{Scott_Goedel}. For Scott,
an essential property of an individual must be possessed by
him/her. For G\"odel, this is not required.

G\"odel's omission has been
considered inessential and merely an oversight by many. 
For more than four decades, its serious consequences remained unnoticed, 
despite numerous analysis and criticisms of the
argument.%  Especially since the discovery by Sobel
% \cite{Sobel} that modal collapse (MC)\footnote{The modal collapse,
%   $\phi\rightarrow \Box \phi$, states that contingent truth implies
%   necessary truth; it can be interpreted as \textit{everything is
%     pre-determined} or even \textit{there is no free will}. } is
% entailed by G\"odel's (or also Scott's) axioms, several variants have
% been proposed
% \cite{Anderson,AndersonGettings,Hajek1,Hajek2,Hajek3,Bjordal}
% attempting to avoid the modal collapse. Many of these variants omit
% the crucial conjunct in the definition of essence as well.\footnote{As
%   these variants also change other axioms, on which the inconsistency
%   of G\"odel's axioms depends, it is not necessarily the case that
%   these variants are also inconsistent; they must be analyzed
%   separately.} Opponents of the argument
% (e.g. Oppy [\cite[p.226--227]{oppy96:_goedel_ontol_argum};
% \cite[p.364]{oppy00:_respon_gettin};
% \cite[p.1068]{oppy08:_higher_order_ontol_argum}]) have also proposed
% parodies and other criticisms, referring to variants where the
% conjunct is omitted.
